Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-04-24 Manuscript/manuscript.Rmd

**Background.** Chromosome conformation capture sequencing technologies have shown that the three-dimensional (3D) structure of the genome is folded into distinct compartments, known as topologically associated domains (TADs) - units of coordinated gene expression. The location of TAD boundaries is highly conserved, suggesting the presence of epigenomic marks aiding in TAD formation. The ability to predict which epigenomic features are most associated with the TAD boundaries will allow for better understanding the regulatory role of the 3D structure of the genome.

@@

The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries. Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. Although several useful insights have been made, such as the convergent binding of CTCF [@ong2014ctcf; @ghirlando2016ctcf], many methods ignore key characteristics of domain data, such as the impact of heavily imbalanced number of boundaries vs. other regions. In a classification setting, when one is attempting to discriminate between regions with and without TAD boundaries, the class imbalance issue will contribute to small imbalance ratios (defined as the ratio of minority to majority classes[@Orriols-Puig2008]). Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class;@chen2013influence;@jeni2013facing]. Consequently, careful consideration must be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

@@

Multiple random forest classification algorithms were built on domain data extracted from Hi-C data for 2 cell lines $C = \{\textit{GM12878}, \textit{K562}\}$, at four different resolutions $R = \{\textit{10 kb}, \textit{25 kb}, \textit{50 kb}, \textit{100 kb\}}$. For each classifier, we compared model performances using 3 separate predictor types $P = \{\textit{OC}, \textit{OP}, \textit{Distance}\}$ and 4 class balancing techniques $B = \{\textit{None}, \textit{RUS}, \textit{ROS}, \textit{SMOTE}\}$. A system produced from the ensemble framework was defined as: $S = \{\textit{c}, \textit{r}, \textit{p}, \textit{b}\}$, where $c \in C$, $r \in R$, $p \in P$, and $b \in B$. Letting $|X|$ denote the cardinality of each set, we evaluated $|C| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $96 (= 2 \times 4 \times 3 \times 4)$ separate models.

@@

To assess the full severity of class imbalance present when constructing the models, we investigated the relationship between the total number of bins (sample observations) included at each resolution and the subsequent imbalance ratio (IR). As expected, the total number of genomic bins decreased as resolution decreased (Figure 4A). Specifically, at lower resolutions, the linear genome was binned into larger units, thereby creating fewer bins. however, the IRs remained relatively stable and consistent across resolutions and cell lines (Figure 4B). We found that there was an average IR of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for the GM12878 and K562 cell lines respectively. Here, the low imbalance ratios indicated a wide disparity between the number of genomic bins that overlap with a flanked TAD boundary and the much larger number that did not.

>>>>

**Background.** Chromosome conformation capture sequencing technologies have shown that the three-dimensional (3D) structure of the genome is folded into distinct compartments, known as topologically associated domains (TADs) - units of coordinated gene expression. The location of TAD boundaries is highly conserved, suggesting the presence of epigenomic marks aiding in TAD formation. [The ability to predict which epigenomic features are most associated with the TAD boundaries will allow for better understanding the regulatory role of the 3D structure of the genome. - ??? Weak]

@@

The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries. Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. Although several useful insights have been made, such as the convergent binding of CTCF [@ong2014ctcf; @ghirlando2016ctcf], [??? Describe other hallmarks of TAD boundaries, with references. And, most importantly, why you are doing your study?]

many methods ignore key characteristics of domain data, such as the impact of heavily imbalanced number of boundaries vs. other regions. In a classification setting, when one is attempting to discriminate between regions with and without TAD boundaries, the class imbalance issue will contribute to small imbalance ratios (defined as the ratio of minority to majority classes[@Orriols-Puig2008]). Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class;@chen2013influence;@jeni2013facing]. Consequently, careful consideration must be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

@@

Multiple random forest classification algorithms [??? What you mean by multiple? Random forest already has multiplicity in it.] were built on domain data extracted from Hi-C data for 2 cell lines $C = \{\textit{GM12878}, \textit{K562}\}$, at four different resolutions $R = \{\textit{10 kb}, \textit{25 kb}, \textit{50 kb}, \textit{100 kb\}}$. [??? Bad spacing, for the whole paragraph. Why you are not using plain and simple Markdown?] For each classifier, we compared model performances using 3 separate predictor types $P = \{\textit{OC}, \textit{OP}, \textit{Distance}\}$ and 4 class balancing techniques $B = \{\textit{None}, \textit{RUS}, \textit{ROS}, \textit{SMOTE}\}$. A system produced from the ensemble framework was defined as: $S = \{\textit{c}, \textit{r}, \textit{p}, \textit{b}\}$, where $c \in C$, $r \in R$, $p \in P$, and $b \in B$. Letting $|X|$ denote the cardinality of each set, we evaluated $|C| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $96 (= 2 \times 4 \times 3 \times 4)$ separate models.

@@

To assess the full [???] severity of class imbalance present when constructing the models, we investigated the relationship between the total number of bins (sample observations) included at each resolution [??? That's mundane, supplemental at best] and the subsequent imbalance ratio (IR). [??? Did you ever defined it?] As expected, the total number of genomic bins decreased as resolution decreased (Figure 4A). Specifically, at lower resolutions, the linear genome was binned into larger units, thereby creating fewer bins. however, the IRs remained relatively stable and consistent across resolutions and cell lines (Figure 4B). We found that there was an average IR of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for the GM12878 and K562 cell lines respectively. Here, the low imbalance ratios indicated a wide disparity between the number of genomic bins that overlap with a flanked TAD boundary and the much larger number that did not.
